---
title: "Using R Markdown (STAT 542)"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
options(width = 1000)
```

## Name: Ruoqing Zhu (rqzhu@illinois.edu)

\section{1. An example: generating data and fitting a linear regression}

\textcolor{red}{First, you need to install required packages if you haven't.} We can then generate some data from a linear model, and solve for $\widehat \beta$.

```{r}
  library(MASS)
  set.seed(1)
  P = 4; N = 200
  V <- 0.5^abs(outer(1:P, 1:P, "-"))
  X = as.matrix(mvrnorm(N, mu=rep(0,P), Sigma=V))
  beta = as.matrix(c(1, 1, 0.5, 0.5))
  Y = X %*% beta + rnorm(N)
  
  # then the estimated beta is 
  hbeta = solve(t(X) %*% X) %*% t(X) %*% Y
  hbeta
```

The above calculation follows the normal equaitons:

$$\widehat \beta = (X'X)^{-1} X'Y.$$

We can also treat this as an optimization problem by minimizing the $\ell_2$ loss function: 

$$\widehat \beta = \arg\min \frac{1}{2}\lVert Y - X\beta \rVert_2^2.$$

This can be done through some numerical optimization methods (\textcolor{red}{although this is entirely NOT necessary and is an overkill...}):

```{r}
  f <- function(b, X, Y)
  {
    0.5*sum((Y - X %*% as.matrix(b))^2)
  }
  
  solution = optim(rep(0, P), f, X = X, Y = Y, method = "BFGS")
  solution$par
```

Many machine learning and statistical learning problems are eventually an optimization problem, although they are much more difficult than solving a linear regression. 

\section{2. Some other things}

When submitting your homework, please **\textcolor{red}{simplify your R output}** so that you don't produce an excessively long report. For example, printing out a data matrix longer than a page is not allowed. Code like the following should have their result hidden by including `results="hide"` rather than displaying the entire vector of $Y$:

```{r, results="hide"}
  Y
```

or this will be better
```{r}
  head(Y, 3)
```

If you really want to show what $Y$ looks like, consider doing a histogram or box plot. \textcolor{red}{Your `R` code should always come with explanations} so that your TA and I can understand your intention. Including a graph can be helpful sometimes, but make sure that you manage the size and layout properly. This can be done by speicfying some parameters when writing the code block (see the .Rmd source file):

```{r, out.width='.49\\linewidth', fig.width = 3, fig.width = 6, fig.show='hold',fig.align='center'}
  plot(X[,1], Y, xlab = "X1", main = "PLOT 1", col = "darkorange", pch = 19)
  plot(X[,3], Y, xlab = "X3", main = "PLOT 2", col = "deepskyblue", pch = 19)
  legend("bottomright", "a sample point", pch = 19, col = "deepskyblue", cex = 1.2)
```

If you need additional help, let your TA or me know. Also, sometimes your best friend is google.

