```{r include = FALSE}
  options(width = 100)
  knitr::opts_knit$set(global.par = TRUE)
  knitr::opts_chunk$set(collapse = TRUE, fig.dim = c(6, 6), 
                        out.width = "45%", fig.align = 'center')
```
```{r include = FALSE}
  par(bg="transparent")
  par(mar=c(2,2,2,2))
```

# Visual Studio Code

## Basics and Resources

Visual Studio Code (VS Code) is another popular IDE for programming in pretty much all languages. With the recent development of GitHub Copilot (X), it makes VS Code a very attractive platform. Here is a place to get things started. Please note that in Windows, running the applications as Administrator is probably needed to install related components. 

  * [Official Overview](https://code.visualstudio.com/docs/languages/r)

To get things working, you only need to

  * Install R
  * Install VS Code
  * Install R Extension in VS Code
  * Install the `languageserver` package `install.packages("languageserver")`

To install `radian` (for interactive R terminal), you need to 

  * Install Python and [pip](https://pip.pypa.io/en/stable/installation/) (if you have Anaconda, that should be included already)
  * Use `pip3 install -U radian` to install `radian`
  * Go to settings in VS Code, and find `r.path`, add your `radian.exe` file path to it
  
