\def\cD{\cal{D}}
\def\bA{\mathbf{A}}
\def\bB{\mathbf{B}}
\def\bX{\mathbf{X}}
\def\bH{\mathbf{H}}
\def\bI{\mathbf{I}}
\def\bU{\mathbf{U}}
\def\bD{\mathbf{D}}
\def\bV{\mathbf{V}}
\def\bS{\mathbf{S}}
\def\bx{\mathbf{x}}
\def\by{\mathbf{y}}
\def\bs{\mathbf{s}}
\def\br{\mathbf{r}}
\def\bu{\mathbf{u}}
\def\be{\mathbf{e}}
\def\bv{\mathbf{v}}
\def\bp{\mathbf{p}}
\def\bzero{\mathbf{0}}
\def\bbeta{\boldsymbol \beta}
\def\bmu{\boldsymbol \mu}
\def\bepsilon{\boldsymbol \epsilon}
\def\T{\text{T}}
\def\Trace{\text{Trace}}
\def\Cov{\text{Cov}}
\def\Var{\text{Var}}
\def\E{\text{E}}
\def\pr{\text{pr}}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

```{r include = FALSE}
  options(width = 100)
  knitr::opts_knit$set(global.par = TRUE)
  knitr::opts_chunk$set(collapse = TRUE, fig.dim = c(6, 6),
                        out.width = "45%", fig.align = 'center')
```
```{r include = FALSE}
  par(bg="transparent")
  par(mar=c(2,2,2,2))
```

# (PART) Nonparametric Models {-}

# Spline
