# (PART) Basics Knowledge {-}

# Introducing R, RStudio and R Markdown

R is a free-to-use software that is very popular in statistical computing. You can download R from its [official website](https://www.r-project.org/). Another software that makes using R easier is RStudio, which is available at [here](https://www.rstudio.com/). You can find many on-line guides that help you to set-up these two software, for example, this [\color{blue} YouTube video](https://www.youtube.com/watch?v=cX532N_XLIs&t=19s/). R Markdown is a build-in feature of RStudio. It works like an integration of LaTex and programming playground that complies source code into nice-looking PDF, html, or MS Word files. This book is created using an extension of R Markdown, developed by [Yihui Xie](https://bookdown.org/yihui/bookdown/). 

## Resources and Guides

There are many online resources for how to use R, RStudio, and R Markdown. For example, David Dalpiaz's other online book [Applied Statistics with R](http://daviddalpiaz.github.io/appliedstats/) contains an introduction to using them. There are also other online documentations such as 

* [Install R and RStudio](https://www.youtube.com/watch?v=cX532N_XLIs&t=19s/)
* [R tutorial](http://www.r-tutor.com/r-introduction)
* [Data in R Playlist (video)](https://www.youtube.com/playlist?list=PLBgxzZMu3GpPojVSoriMTWQCUno_3hjNi)
* [R and RStudio Playlist (video)](https://www.youtube.com/playlist?list=PLBgxzZMu3GpMjYhX7jLm5B9gEV7AOOJ5w)
* [R Markdown Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
* [R Markdown Playlist (video)](https://www.youtube.com/playlist?list=PLBgxzZMu3GpNgd07DwmS-2odHtMO6MWGH)

It is worth to mention that once you become a developer of R packages using `C/C++` (add-on of R for performing specific tasks), and you also happen to use Windows like I do, you have to install this [Rtools](https://cran.r-project.org/bin/windows/Rtools/) that contains compilers. This is also needed if you want to mannually install any R package using a "source" (`.tar.gz` files) instead of using the so-called "binaries" (`.zip` files). 

## Demonstrating Examples

We will briefly cover some basic R calculations and operations. 

### Basic Mathematical Operations

Try type-in the following commands into your R console and start to explore yourself. Most of them are self-explanatory. Lines with a `#` in the front are comments, which will not be executed. Lines with `##` in the front are outputs. 

```{r, collapse=TRUE}
# Basic mathematical operations
  1 + 3
  3*5
  3^5
  exp(2)
  log(3)
  log2(3)
  factorial(5)
```

### Data Objects

Data objects can be a complicated topic for people who never used R before. Most common data objects are `vector`, `matrix`, `list`, and `data.frame'. Operations on vectors are matrices are faily intuitive.

```{r, collapse=TRUE}
# creating a vector
  c(1,2,3,4)
# creating matrix from a vector
  matrix(c(1,2,3,4), 2, 2)
  x = c(1,1,1,0,0,0); y = c(1,0,1,0,1,0)
  cbind(x,y)
# matrix multiplication using '%*%'
  matrix(c(1,2,3,4), 2, 2) %*% t(cbind(x,y))
```  

Simple mathematical operations on vectors and matrices are usually element-wise. You can easliy extract certain elements of them by using the `[]` operator, like a `C` programming reference style. 

```{r, collapse=TRUE}  
# some simple operations 
  x[3]
  x[2:5]
  cbind(x,y)[1:2, ]
  (x + y)^2
  length(x)
  dim(cbind(x,y))
# A warning will be issued when R detects something wrong. Results may still be produced.
  x + c(1,2,3,4)
```

`list()` simply creates a list of objects (of any type). However, some operators cannot be directly applied to a list in a similar way as to vectors or matrices. Model fitting results in R are usually stored as a list. 

```{r, collapse=TRUE}  
# creating a list
  x = list(c(1,2), "hello", matrix(c(1,2,3,4), 2, 2))
  x[[1]]
```



\section{3. Random Number Generations and Descriptive Statistics}

Random number generation is important for statistical simulation. 
```{r, collapse=TRUE}
    # generate 10 Bernoulli random variables as a vector
    rbinom(n=10, size = 1, prob = 0.5)
    # 2 random normally distributed variables
    rnorm(n=4, mean = 1, sd = 2)

```
\vspace{12pt}
If we need to replicate the results, we can set random seed
```{r, collapse=TRUE}
    # after setting the seed, the two runs will generate exactly the same "random" numbers
    set.seed(1)
    rnorm(n=4, mean = 1, sd = 2)
    set.seed(1)
    rnorm(n=4, mean = 1, sd = 2)
```
\vspace{12pt}
Statistical functions that provides a summary of the data
```{r, collapse=TRUE}
    x = rnorm(n=100, mean = 1, sd = 2)
    y = rnorm(n=100, mean = 2, sd = 1)
    sum(x)
    mean(x)
    var(x)
    median(x)
    quantile(x, c(0.25, 0.5, 0.75))
    cor(x, y)
```
\vspace{12pt}
For discrete data, we usually use the table function
```{r, collapse=TRUE}
    set.seed(1); n = 1000
    x = rbinom(n, size = 1, prob = 0.75)
    y = rbinom(n, size = 3, prob = c(0.4, 0.3, 0.2, 0.1))
    table(x)
    table(x, y)
```
For a mixture of discrete and continuous data (multiple variables), we often use a data.frame
```{r, collapse = TRUE}
    # data.frame is a special data structure that can store both factor and numeric variables
    z = runif(n, min = 18, max = 65)
    data = data.frame("Gender" = as.factor(x), "Group" = as.factor(y), "Age" = z)
    levels(data$Gender) = c("male", "female")
    levels(data$Group) = c("patient", "physician", "engineer", "statistician")
    
    # a peek at the top 3 entries of the data
    head(data, 3)
    
    # a brief summary
    summary(data)
    
    # generate a 2 by 2 table that summarizes Gender and Group
    table(data[, 1:2])
```

\section{4. Simple Statistical Tests}

Fisher Exact Test and Chi-square test are tests of independence between two nominal variables.

```{r, collapse=TRUE}
    # the test p-value is not significant
    # recall that we generated the two variables independently.
    fisher.test(table(data[, 1:2]))

    # chi-square test
    chisq.test(table(data[, 1:2]))
```

For continuous variables, we can calculate and test Pearson's correlation, Spearman's rho or Kendall' tau
```{r, collapse=TRUE}
    set.seed(1); n = 30
    x = rnorm(n)
    y = x + rnorm(n, sd = 2)
    z = x + rnorm(n, sd = 2)
    
    # one can specify method = "kendall" or "spearman" to perform other correlations or tests
    cor(y, z)
    cor.test(y, z)
```

\section{5. Regression Models}

A simple linear regression assumes the underlying model $Y = \beta X + \epsilon$. With observed data, we can estimate the regression coefficients:
```{r, collapse=TRUE}
    # the lm() function is the most commonly used
    fit = lm(y~x)
    summary(fit)
```

\section{6. Data Visualization}
A graphical view usually helps understand the data better. There are a variety of ways to customize your plot, such as color and shape.

```{r, out.width='.49\\linewidth', fig.width = 3, fig.width = 6, fig.show='hold',fig.align='center'}
    plot(x, y, xlab = "x", ylab = "y", main = "A plot", col = "darkorange", pch = 19)
    abline(a = fit$coefficients[1], b = fit$coefficients[2], 
           col = "deepskyblue", lty = 2, lwd = 2)
    legend("topleft", c("observed points", "fitted line"), lty = c(NA, 2), 
           pch = c(19, NA), col = c("darkorange", "deepskyblue"))
```

\section{7. Other Things}

\subsection{7.1 Read-in and Save Data}
R can read-in data from many different sources such as \texttt{.txt}, \texttt{.csv}, etc. For example, \texttt{read.csv()} can be used to import \texttt{.csv} files. The first argument should be specified as the path to the data file, or just the name of the file if the current working directory is the same as the data file. R objects, especially matrices, can be saved into these standard files. Use functions such as \texttt{write.table()} and \texttt{write.csv} to perform this. We can also save any object into \texttt{.RData} file, which can be loaded later on. To do this try functions \texttt{save.image()} and \texttt{save()}.

\subsection{7.2 R Packages}
One of the most important features of R is its massive collection of packages. A package is like an add-on that can be downloaded and installed and perform additional function and analysis. 
```{r, collapse=TRUE}
    # The MASS package can be used to generate multivariate normal distribution 
    library(MASS)
    P = 4; N = 200
    V <- 0.5^abs(outer(1:P, 1:P, "-"))
    X = as.matrix(mvrnorm(N, mu=rep(0,P), Sigma=V))
    head(X, 3)
```

\subsection{7.3 Get Helps}
To get reference about a particular function, one can put a question mark in front of a function name to see details: 
```{r, collapse=TRUE, eval=FALSE}
    # This will open up the web browser for the on-line document 
    ?mvrnorm
    ?save.image
```









