# (PART) Basics Knowledge {-}

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 1000)
```

# R and RStudio

The first step is to download and install [R](https://www.r-project.org/) and [RStudio](https://www.rstudio.com/products/rstudio/download/#download). Most steps should be self-explanatory. You can also find many online guides for step-by-step instruction, such as [this YouTube video](https://www.youtube.com/watch?v=cX532N_XLIs&t=19s/). However, be aware that some details may have been changed over the years. 

After installing both, open your RStudio, you should see four panes: 

  * Source pane on top-left where you write code in to files
  * Console on bottom-left where the code is inputted into R
  * Environment (and other tabs) on top-right where you can see current variables and objects you defined
  * File (and other tabs) on bottom-right which is essentially a file borrower

We will mainly use the left two panes. You can either directly input code into the console to run for results, or edit your code in a file and run them in chunks or as a whole. 

## Resources and Guides  {#r-basic}

There are many online resources for how to use R, RStudio. For example, David Dalpiaz's other online book [Applied Statistics with R](http://daviddalpiaz.github.io/appliedstats/) contains an introduction to using them. There are also other online documentation such as 

* [Install R and RStudio](https://www.youtube.com/watch?v=cX532N_XLIs&t=19s/)
* [R tutorial](http://www.r-tutor.com/r-introduction)
* [Data in R Play-list (video)](https://www.youtube.com/playlist?list=PLBgxzZMu3GpPojVSoriMTWQCUno_3hjNi)
* [R and RStudio Play-list (video)](https://www.youtube.com/playlist?list=PLBgxzZMu3GpMjYhX7jLm5B9gEV7AOOJ5w)

It is worth to mention that once you become an advanced user, and possibly a developer of R packages using `C/C++` (add-on of R for performing specific tasks), and you also happen to use Windows like I do, you will have to install [Rtools](https://cran.r-project.org/bin/windows/Rtools/) that contains the gcc compilers. This is also needed if you want to install any R package from a "source" (`.tar.gz`) file instead of using the so-called "binaries" (`.zip` files). 

## Basic Mathematical Operations

Basic R calculations and operations should be self-explanatory. Try to type-in the following commands into your R console and start to explore yourself. Lines with a `#` in the front are comments, which will not be executed. Lines with `##` in the front are outputs you should expect. 

```{r, collapse=TRUE}
    # Basic mathematical operations
    1 + 3
    1 - 3
    1 * 3
    1 / 3
    3^5
    4^(-1/2)
    pi
    
    # some math functions
    sqrt(4)
    exp(1)
    log(3)
    log2(16)
    log(15, base = 3)
    factorial(5)
    sin(pi)
```

If you want to see more information about a particular function or operator in R, the easiest way is to get the reference document. Put a question mark in front of a function name:

```{r, collapse=TRUE, eval=FALSE}
    # In a default R console window, this will open up a web browser.
    # In RStudio, this will be displayed at the ‘Help’ window at the bottom-right penal (Help tab). 
    ?log10
    ?cos
```

## Data Objects

Data objects can be a complicated topic for people who never used R before. The most common data objects are `vector`, `matrix`, `list`, and `data.frame`. They are defined using a specific syntax. To define a vector, we use `c` followed by `()`, where the elements within the parenthesis are separated using comma. You can save the vector and name as something else. For example

```{r, collapse=TRUE}
    # creating a vector
    c(1,2,3,4)
    c("a", "b", "c")
    
    # define a new vector object, called `x`
    x = c(1,1,1,0,0,0)
```

After defining this object `x`, it should also appear on your top-right environment pane. To access elements in an object, we use the `[]` operator, like a `C` programming reference style.

```{r, collapse=TRUE}
    # getting the second element in x
    x[2]

    # getting the second to the fourth element in x
    x[2:4]
```

Similarly, we can create and access matrices.

```{r, collapse=TRUE}
    # create a matrix by providing all of its elements
    # the elements are filled to the matrix by column
    matrix(c(1,2,3,4), 2, 2)

    # create a matrix by column-bind vectors
    y = c(1,0,1,0,1,0)
    cbind(x, y)

    # access elements in a matrix
    # Note that in R, upper and lower cases are treated as two different objects
    X = matrix(c(1:16), 4, 4)
    X
    X[2, 3]
    X[1, ]
    
    # getting a sub-matrix of X
    X[1:2, 3:4]
```  

Simple mathematical operations on vectors and matrices are usually element-wise. For matrix multiplications, you should use `%*%`.

```{r, collapse=TRUE}  
    # adding two vectors
    (x + y)^2

    # getting the length of a vector
    length(x)
    
    # matrix multiplication
    X %*% X
    
    # getting the dimension of a matrix
    dim(X)
    
    # A warning will be issued when R detects something wrong
    # Results may still be produced however
    y + c(1,2,3,4)
```

`list()` creates a list of objects (of any type). However, some operators cannot be directly applied to a list in a similar way as to vectors or matrices. Model fitting results in R are usually stored as a list (for example, the `lm()` function used in Chapter \@ref(linear-reg).

```{r, collapse=TRUE}  
    # creating a list
    x = list(c(1,2), "hello", matrix(c(1,2,3,4), 2, 2))

    # accessing its elements using double brackets `[[]]` 
    x[[1]]
```

`data.frame()` creates a list of vectors of equal length, and display them as a matrix-like object where each vector is a column of the matrix. It is mainly used for storing data. This will be our most frequently used data object. For example, in the famous `iris` data, the first four columns are numerical variables, while the last column is a categorical variable with three levels: `setosa`, `versicolor`, and `virginica`:

```{r, collapse=TRUE}  
    # The iris data is included with base R, so we can use them directly
    # This will create a copy of the data into your environment
    data(iris)

    # the head function peeks the first several rows of the dataset 
    head(iris, n = 3)
    
    # each column usually contains a column (variable) name 
    colnames(iris)
    
    # data frame can be called by each individual column, which will be a vector
    iris$Species
    iris$Species[2:4]
    
    # the summary function can be used to view summary statistics of all variables
    summary(iris)
```

## Read-in Data from Other Sources

Data can be imported from a variety of sources. More commonly, a dataset can be stored in `.txt` and `.csv` files. 

```{r, collapse=TRUE}  
    # read-in data
    birthrate = read.csv("data/birthrate.csv")
    head(birthrate)
    # to see how many observations (rows) and variables (columns) in a dataset
    dim(birthrate)
```

R data can also be saved into other formats. The more efficient way, assuming that you are going to load these file back to R in the future, is to save them as `.RData` file. Usually, for a large dataset, this reduces the time spend on reading in the data. 
```{r, collapse=TRUE, eval = FALSE}
    # saving a object to .RData file
    save(birthrate, file = "mydata.RData")
    # you can specify multiple objects too
    save(birthrate, iris, file = "mydata.RData")
    # load the data again
    load("mydata.RData")
    # save data to .csv file
    write.csv(birthrate, file = "mydata.csv")
```

## Using Packages

Packages are written and contributed to R by individuals. They provide additional features (functions or data) that serve particular needs. The package `ElemStatLearn` is created for the textbook [@hastie2001elements] and contains may popular datasets. A package needs to be installed and loaded to your local computer. 

```{r, collapse=TRUE, eval=FALSE}
    # install packages
    install.packages("ElemStatLearn")
    # load package
    library(ElemStatLearn)
```

```{r, echo=FALSE}
    library(ElemStatLearn)
```

```{r, collapse=TRUE}  
    # load a dataset from the package 
    data(SAheart)
    head(SAheart)
```

## Explore Yourself

There is no guide that can exhaust all information. I found the best learning approach is to perform a specific task and google your way to the success. Oftentimes, Stack Overflow is my best friend, especially when I am developing new packages. Also, read the reference manual carefully if you use a particular package or function. A reference manual (for packages on CRAN) can always be found at [https://cran.r-project.org/web/packages/package_name](https://cran.r-project.org/web/packages/package_name). 