# Basics of Probability and Statistics

## Random Number Generation 

Random number generation is important for statistical simulation. R provides random number generators for many commonly used distributions, such as binomial (`rbinom`), normal (`rnorm`), t (`rt`) etc. The syntax is usually a letter "r" followed by the name of the distribution. 

```{r, collapse=TRUE}
    # generate 10 independent Bernoulli random variables as a vector
    rbinom(n=10, size = 1, prob = 0.5)
    # 4 independent random standard normal variables
    rnorm(n=4)
```

Setting the seed before generating random numbers will allow us to replicate the results when necessary. 
```{r, collapse=TRUE}
    # after setting the seed, the two runs will generate exactly the same "random" numbers
    set.seed(1)
    rnorm(n=4, mean = 1, sd = 2)
    set.seed(1)
    rnorm(n=4, mean = 1, sd = 2)
```

Some more complicated distributions require additional package. For example, the `MASS` package can be used to generate multivariate normal distribution. One needs to specify a covariance matrix. 
```{r, collapse=TRUE}
    library(MASS)
    P = 4
    V <- 0.5^abs(outer(1:P, 1:P, "-"))
    mvrnorm(3, mu=1:P, Sigma=V)
```

## Summary Statistics and Data Visualization

Statistical functions that provides a summary of the data
```{r, collapse=TRUE}
    x = rnorm(n=100, mean = 1, sd = 2)
    y = rnorm(n=100, mean = 2, sd = 1)
    sum(x)
    mean(x)
    var(x)
    median(x)
    quantile(x, c(0.25, 0.5, 0.75))
    cor(x, y)
```

For discrete data, we can use the table function
```{r, collapse=TRUE}
    library(ElemStatLearn)
    data(SAheart)
    table(SAheart$famhist)
    table(SAheart[, c("famhist","chd")])
```

Fisher's exact test and the Chi-square test are tests of independence between two nominal variables.

```{r, collapse=TRUE}
    # We can test the association between family history (famhist) and 
    # the indicator of coronary heart disease (chd)
    # using Fisher's Exact
    fisher.test(table(SAheart[, c("famhist","chd")]))

    # or the Chi-square test
    chisq.test(table(SAheart[, c("famhist","chd")]))
```

For continuous variables, data visualization can be very helpful. There are many different ways to customize a plot, such as changing the color, shape, label, etc. For more advanced features, the R package `ggplot2` is a very popular choice. 
```{r, collapse=TRUE, fig.width=6, fig.height=4}
    # We use the birthrate data introduced earlier for this example
    birthrate = read.csv("data/birthrate.csv")
    plot(birthrate, pch = 19, col = "darkorange", ylab = "Birth Rate",
         main = "U.S. birth rate (1917 - 2003, per 10000)")
```

Correlations and correlation plots can be used to summarize more varaibles. However, becareful that factors may not be supported by this feature and could cause errors. 
```{r, collapse=TRUE, fig.width=5, fig.height=5}
    # load the package with loading message suppressed
    suppressMessages(library(PerformanceAnalytics))
    chart.Correlation(SAheart[, c(1:3)], histogram=TRUE, pch="+")
```

3-dimentional plots are also alternative ways to visualize data. We demonstrate an example using the `plot3D` package and the `scatter3D` function. The package `rgl` can allow for interactive plot with rotating and zooming. 

```{r, collapse=TRUE, fig.width=6, fig.height=6}
    library(plot3D)
    scatter3D(SAheart$ldl, SAheart$age, log(1+SAheart$tobacco), 
              xlab = "LDL", ylab = "Age", zlab = "Tobacco",
              pch = 18, bty = "u", col.var = SAheart$chd, 
              col = ifelse(SAheart$chd == 1, "darkorange", "deepskyblue"),
              colkey = FALSE)
```