# Optimizations in `R'

**TODO:** Instead of specifically considering regression, change the focus of this chapter to modeling, with regression as an example.

This chapter will recap the basics of performing regression analyses in `R`. For more detailed coverage, see [Applied Statistics with `R`](http://daviddalpiaz.github.io/appliedstats/).

We will use the [Advertising data](data/Advertising.csv) associated with [Introduction to Statistical Learning](http://www-bcf.usc.edu/~gareth/ISL/data.html).

```{r, message = FALSE, warning = FALSE}
library(readr)
Advertising = read_csv("data/Advertising.csv")
```

After loading data into `R`, our first step should **always** be to inspect the data. We will start by simply printing some observations in order to understand the basic structure of the data.

```{r}
Advertising
```

Because the data was read using `read_csv()`, `Advertising` is a tibble. We see that there are a total of `r nrow(Advertising)` observations and `r ncol(Advertising)` variables, each of which is numeric. (Specifically double-precision vectors, but more importantly they are numbers.) For the purpose of this analysis, `Sales` will be the **response variable**. That is, we seek to understand the relationship between `Sales`, and the **predictor variables**: `TV`, `Radio`, and `Newspaper`.

## A simple optimization problem

After investigating the structure of the data, the next step should be to visualize the data. Since we have only numeric variables, we should consider **scatter plots**.

