---
output:
  html_document: default
  pdf_document: default
---

# R Basics

## Resources and Guides

There are many online resources for how to use R, RStudio, and R Markdown. For example, David Dalpiaz's other online book [Applied Statistics with R](http://daviddalpiaz.github.io/appliedstats/) contains an introduction to using them. There are also other online documentation such as 

* [Install R and RStudio](https://www.youtube.com/watch?v=cX532N_XLIs&t=19s/)
* [R tutorial](http://www.r-tutor.com/r-introduction)
* [Data in R Play-list (video)](https://www.youtube.com/playlist?list=PLBgxzZMu3GpPojVSoriMTWQCUno_3hjNi)
* [R and RStudio Play-list (video)](https://www.youtube.com/playlist?list=PLBgxzZMu3GpMjYhX7jLm5B9gEV7AOOJ5w)
* [R Markdown Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
* [R Markdown Play-list (video)](https://www.youtube.com/playlist?list=PLBgxzZMu3GpNgd07DwmS-2odHtMO6MWGH)

It is worth to mention that once you become a developer of R packages using `C/C++` (add-on of R for performing specific tasks), and you also happen to use Windows like I do, you have to install this [Rtools](https://cran.r-project.org/bin/windows/Rtools/) that contains compilers. This is also needed if you want to manually install any R package using a "source" (`.tar.gz` files) instead of using the so-called "binaries" (`.zip` files). 

## Basic Mathematical Operations

We will briefly cover some basic R calculations and operations. If you want to see more information about a particular function or operator in R, the easiest way is to get the reference document. Put a question mark in front of a function name:

```{r, collapse=TRUE, eval=FALSE}
    # In a default R console window, this will open up a web browser.
    # In RStudio, this will be displayed at the ‘Help’ window at the bottom-right penal. 
    ?log2
    ?matrix
```

Try type-in the following commands into your R console and start to explore yourself. Most of them are self-explanatory. Lines with a `#` in the front are comments, which will not be executed. Lines with `##` in the front are outputs. 

```{r, collapse=TRUE}
    # Basic mathematical operations
    1 + 3
    3*5
    3^5
    exp(2)
    log(3)
    log2(3)
    factorial(5)
```

## Data Objects

Data objects can be a complicated topic for people who never used R before. Most common data objects are `vector`, `matrix`, `list`, and `data.frame`. Operations on vectors are matrices are fairly intuitive.

```{r, collapse=TRUE}
    # creating a vector
    c(1,2,3,4)
    c("a", "b", "c")
    # creating matrix from a vector
    matrix(c(1,2,3,4), 2, 2)
    x = c(1,1,1,0,0,0); y = c(1,0,1,0,1,0)
    cbind(x,y)
    # matrix multiplication using '%*%'
    matrix(c(1,2,3,4), 2, 2) %*% t(cbind(x,y))
```  

Simple mathematical operations on vectors and matrices are usually element-wise. You can easily extract certain elements of them by using the `[]` operator, like a `C` programming reference style. 

```{r, collapse=TRUE}  
    # some simple operations 
    x[3]
    x[2:5]
    cbind(x,y)[1:2, ]
    (x + y)^2
    length(x)
    dim(cbind(x,y))
    # A warning will be issued when R detects something wrong. Results may still be produced.
    x + c(1,2,3,4)
```

`list()` simply creates a list of objects (of any type). However, some operators cannot be directly applied to a list in a similar way as to vectors or matrices. Model fitting results in R are usually stored as a list (for example, the `lm()` function used in Section \@ref(sec:lm).

```{r, collapse=TRUE}  
    # creating a list
    x = list(c(1,2), "hello", matrix(c(1,2,3,4), 2, 2))
    x[[1]]
```

`data.frame()` creates a list of vectors of equal length, and display them as a matrix-like object where each vector is a column of the matrix. It is mainly used for storing data. For example the famous iris data. The first four columns are numerical variables, while the last column is a categorical variable with three levels: `setosa`, `versicolor`, and `virginica`.

```{r, collapse=TRUE}  
    # the head function peeks the first several rows of the dataset 
    head(iris, n = 3)
    # data frame can be called by each individual column
    iris$Species[c(1, 51, 101)]
    # the summary function can be used to view all variables
    summary(iris)
```

## Read-in Data from Other Sources

Data can be imported from a variety of sources. More commonly, a dataset can be stored in `.txt`, `.csv` or other file formats. 
```{r, collapse=TRUE}  
    # read-in data
    birthrate = read.csv("data/birthrate.csv")
    head(birthrate)
    # to see how many observations (rows) and variables (columns) in a dataset
    dim(birthrate)
```

R data can also be saved into other formats. The more efficient way, assuming that you are going to load these file back to R in the future, is to save them as `.RData` file. Usually, for a large dataset, this reduces the time spend on reading in the data. 
```{r, collapse=TRUE, eval = FALSE}  
    # saving a object to .RData file
    save(birthrate, file = "mydata.RData")
    # you can specify multiple objects too
    save(birthrate, iris, file = "mydata.RData")
    # load the data again
    load("mydata.RData")
    # save data to .csv file
    write.csv(birthrate, file = "mydata.csv")
```

## Using Packages

Packages are written and contributed to R by individuals. They provide additional features (functions or data) that serve particular needs. The package `ElemStatLearn` is created for the textbook [@hastie2001elements] and contains may popular datasets. A package needs to be installed and loaded to your local computer. 

```{r, collapse=TRUE, eval=FALSE}
    # install packages
    install.packages("ElemStatLearn")
    # load package
    library(ElemStatLearn)
```

```{r, echo=FALSE}
    library(ElemStatLearn)
```

```{r, collapse=TRUE}  
    # load a dataset from the package 
    data(SAheart)
    head(SAheart)
```

## Explore Yourself

There is no guide that can exhaust all information. I found the best learning approach is to perform a specific task and google your way to the success. Oftentimes, Stack Overflow is my best friend, especially when I am developing new packages. Also, read the reference manual carefully if you use a particular package or function. A reference manual (for packages on CRAN) can always be found at [https://cran.r-project.org/web/packages/package_name](https://cran.r-project.org/web/packages/package_name). 